User guide -- microxanox R Package for microbial oxic and anoxic ecosystem simulations

### 2021-04-16 various_stuff/dev_vignettes/User-guide.Rmd
changed framework
defaults now only refer to the values of the run_simultion arguments used if none are given in the function call

@@

The aims here are:

* Reproduce the results of [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x).
* Add an investigation of how changes in tolerances alter the response to environmental change.

We also add:

* Option to include random noise in substrate concentrations.
* Option to temporally vary oxygen diffusivity.
* Option to include immigration.
* Option to include a minimum abundances.

Here we use only the published (Bush et al 2017) dynamical model, in which there is not biodiversity within each of the three functional groups. That is, there is only one parameter set per functional group.

Here is the original publication:
https://www.nature.com/articles/s41467-017-00912-x

@@

Note that we here use the notation from the table S1 of the supplementary information, and not the notation in the main text. (There is only one small difference between these: In the main text the notation is simplified to y_S_PB (which is y_SR_PB ($y_{SR,PB}$) in Table S1) and y_S_SB (which is y_SO_SB ($y_{SO,SB}$) in Table S1).

# Methods

## Modelling framework

* Define the default simulation conditions.
* Use these defaults unless other conditions are specified.
* The function call `run_simulation()` would therefore run a simulation with default conditions.
* Whereas, for example, the function call `run_simulation(initial_state = init_state_favouring_anoxic)` would run the simulation with the specified initial state values.

## Dynamic model

There are two versions...

The ordinary differential equations for the rates of change are specified in the function `bushplus_dynamic_model_strains()`. This is an augmented version that can handle multiple strains within each of the three functional groups. The basic version that has only one strain per functional group is `bushplus_dynamic_model()`.

@@

>>>>

```{r setup, echo = FALSE, output = FALSE, message = FALSE}

knitr::opts_chunk$set(cache = FALSE,
                      fig.align = "centre", fig.width = 7, fig.height = 7)

library(tidyverse)
library(deSolve)
library(rootSolve)
library(here)
library(microxanox)
library(patchwork)
```

```{r echo = TRUE}
library(microxanox)
```

@@

The aims of the `microxanox` package are:

* Facilitate reproduction of the results of [Bush et al 2017](https://www.nature.com/articles/s41467-017-00912-x).
* Include new functionality that allows new research questions to be answered.

We added functionality for:

* adding random noise in substrate concentrations.
* adding temporally vary oxygen diffusivity.
* including immigration.
* setting minimum population abundances
* changing organismal tolerances to environmental conditions.
* Changing the number of strains per functional group.

Here we use only the published (Bush et al 2017) dynamical model, in which there is not biodiversity within each of the three functional groups. That is, there is only one parameter set per functional group. Please see the vignette for multi-strain simulations for information about implementing those.

Here (again) is the original publication:
https://www.nature.com/articles/s41467-017-00912-x

@@

Note that om this package we use the notation from the table S1 of the supplementary information, and not the notation in the main text. (There is only one small difference between these: In the main text the notation is simplified to y_S_PB (which is y_SR_PB ($y_{SR,PB}$) in Table S1) and y_S_SB (which is y_SO_SB ($y_{SO,SB}$) in Table S1).

@@

# Modelling framework

* Define the default simulation conditions.
* Use these defaults unless other conditions are specified.
* The function call `run_simulation()` would therefore run a simulation with default conditions.
* The function call `run_simulation(initial_state = init_state_favouring_anoxic)` would run the simulation with the specified initial state values.

# (Default) conditions that must be defined

Conditions that must be set, with an example of setting them:

## The dynamic model

The ordinary differential equations for the rates of change are specified in the function `bushplus_dynamic_model()`. An augmented version `bushplus_dynamic_model_strains()` can handle multiple strains within each of the three functional groups. The multiple strain version can be run with one strain per functional group, and then gives the same results as the basic version. The multiple strain versions is, however, quite a bit slower, even for one strain.

Here we define the default dynamic model to be the basic, one strain per functional group model

```{r}
default_dynamic_model <- bushplus_dynamic_model
```

## The parameter values

The parameter values:

```{r}
default_parameter_values <- return_bushetal2017_parameter_values()
```

## The initial values of state variables:

```{r}
default_initial_state <- return_anoxic_1strain_initial_state()
```

## The duration of the simulation:

```{r}
default_sim_duration <- 2000
```

## The sampling interval:

```{r}
default_sim_sample_interval <- 1
```

## The oxygen diffusivity dynamics

Specified by giving a vector of values of $a_0$. These are then automatically spaced equally along the duration of the simulation. Here we specify no forcing of $a_0$.

```{r}
default_log10a_series <- c(log10(default_parameter_values["a_O"]),
                           log10(default_parameter_values["a_O"]))
```

## Events

These are "things" that happen during the simulation. One has to specify what happens, and when it happens.

#### What happens

Default is to ensure abundances are not below minimum, and to add the amount of noise specified below (none in this vignette).

At present there are two functions available, one for single strain simulations, and one for multi-strain simulations. Both can add noise to a state variable, and/or assure minimum abundance is not exceeded.

```{r}
default_event_definition <- event_definition_1
```

#### Default event interval

Here set to an event every 10 time units.

```{r}
default_event_interval <- 10
```

#### Default noise event options

Default to no noise added to substrate concentrations:

```{r}
default_noise_sigma <- 0
```

#### Default minimum abundance event

Default to following minimum abundances:

```{r}
default_minimum_abundances <- rep(1e2, 3)
names(default_minimum_abundances) <- c("CB", "PB", "SB") 
```

# Warning: Assignment to global environment

Please note that the following objects are created during simulations and are assign to the global environment:

* `log10a_forcing_func`
* `noise_sigma`
* `minimum_abundances`

# Run a simulation

Once all the conditions are assigned to the default objects, then the `run_simulation` function will pick them up without being explicitly given them:

```{r}
default_sim_results <- run_simulation()
```

# Plot the results of a simulation

```{r}
plot_dynamics(default_sim_results)
```

# Run simulation with different defaults

```{r}
initial_state_favouring_oxic <- c(N_CB = 1e7,
           N_PB = 1e5,
           N_SB = 1e5,
           SO = 500,
           SR = 50,
           O = 30,
           P = 10)

new_sim_results <- run_simulation(initial_state = initial_state_favouring_oxic)
plot_dynamics(new_sim_results)
```

# Appendix

The code below was useful for debugging.

```{r eval = FALSE}
dynamic_model = bushplus_dynamic_model
event_definition = default_event_definition
parameter_values = default_parameter_values
event_interval = default_event_interval
noise_sigma = default_noise_sigma
minimum_abundances = default_minimum_abundances
sim_duration = default_sim_duration
sim_sample_interval = default_sim_sample_interval
log10a_series = default_log10a_series
initial_state = default_initial_state
solver_method = "radau"

times <- seq(0,
             sim_duration,
             by = sim_sample_interval)

event_times <- seq(min(times),
                   max(times),
                   by = event_interval)

log10a_forcing <- matrix(ncol=2,
                         byrow=F,
                         data=c(ceiling(seq(0,
                                            max(times),
                                            length=length(log10a_series))),
                                log10a_series))

log10a_forcing <- approxfun(x = log10a_forcing[,1],
                            y = log10a_forcing[,2],
                            method = "linear",
                            rule = 2)
assign("log10a_forcing_func", log10a_forcing, envir = .GlobalEnv)

out <- as.data.frame(ode(y = initial_state,
                         times = times,
                         func = dynamic_model,
                         parms = parameter_values,
                         method = solver_method,
                         events = list(func=event_definition,
                                       time=event_times)))

bushplus_dynamic_model(1, initial_state, default_parameter_values)

sim_res <- run_simulation(dynamic_model = bushplus_dynamic_model,
                           event_definition = default_event_definition,
                           parameter_values = default_parameter_values,
                           event_interval = default_event_interval,
                           noise_sigma = default_noise_sigma,
                           minimum_abundances = default_minimum_abundances,
                           sim_duration = default_sim_duration,
                           sim_sample_interval = default_sim_sample_interval,
                           log10a_series = default_log10a_series,
                           initial_state = default_initial_state,
                           solver_method = "radau")

sim_res <- run_simulation(dynamic_model = bushplus_dynamic_model)

plot_dynamics(sim_res)

```
